---
title: "Portland Crashes"
author: "Thomas Sato"
date: "2022-10-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This dataset is from RLIS, and it contains 321,829 car crashes in the Portland Metro area from 2007 to 2020.

Reading in the data:
```{r}
library(tidyverse)
crashes <- read_csv("Crashes.csv")
orpop <- read_csv("https://raw.githubusercontent.com/thomascsato/Portland-Metro-Area-Car-Crashes/main/Oregon%20Cities%20Population.csv", col_names = c("Rank", "City", "Population"))
Streets <- read_csv("https://raw.githubusercontent.com/thomascsato/Portland-Metro-Area-Car-Crashes/main/Streets.csv")
```


# Distributions

```{r}
ggplot(crashes, aes(x = fct_rev(fct_infreq(GIS_JURIS)))) +
  geom_bar(fill = "skyblue", color = "black") +
  geom_text(aes(label = ..count..), stat = "count", size = 2.5, hjust = 0) +
  coord_flip() +
  ggtitle("Distribution of Car Crashes by City in Portland Metro Area (2007-2020)") +
  theme(plot.title = element_text(hjust = 1.5)) +
  ylab("Number of Car Crashes") +
  xlab("City/Area") +
  scale_y_continuous(limits = c(0, 200000))
```

```{r}
ncrash <- crashes %>%
  group_by(GIS_JURIS) %>%
  summarize(n = n())

# Fairview city population data was inaccurate
orpop[which(orpop$City == "Fairview"), 3] <- 9382

percap <- left_join(ncrash, orpop, by = c("GIS_JURIS" = "City")) %>%
  mutate(cpercap = n / Population) %>%
  arrange(desc(cpercap))

ggplot(percap, aes(fct_reorder(GIS_JURIS, cpercap), cpercap)) +
  geom_col(fill = "chartreuse", color = "black") +
  coord_flip() +
  xlab("City/Area") +
  ylab("Number of Crashes Per Capita") +
  ggtitle("Number of Car Crashes Per Capita in Portland Metro Area From 2007-2020") +
  scale_y_continuous(limits = c(0, 0.4)) +
  theme(plot.title = element_text(hjust = 1.2)) +
  geom_text(aes(label = round(cpercap, 3)), size = 3, hjust = -0.1)
```

# Streets Analysis

Initial Data Transformation
```{r}
# Getting the sum length of all the streets because the vector shapefile has every street in segments
streetlen <- Streets %>%
  group_by(PREFIX, STREETNAME, FTYPE) %>%
  summarize(stlength = sum(LENGTH)) %>%
  # Ifelse statement used in order to not have NAs in some full street names
  mutate(FullStName = ifelse(is.na(PREFIX), paste(STREETNAME, FTYPE, sep = " "),
                             paste(PREFIX, STREETNAME, FTYPE, sep = " ")))

# Dataset containing the number of crashes per street
crashesperstreet <- crashes %>%
  group_by(ST_FULL_NM) %>%
  summarize(n = n()) %>%
  arrange(desc(n))

# Joins together the dataset in order to have a variable for crashes per foot of street
crashesperft <- left_join(crashesperstreet, streetlen, by = c("ST_FULL_NM" = "FullStName")) %>%
  mutate(crashes_per_foot = n / stlength) %>%
  arrange(desc(crashes_per_foot)) %>%
  # Ranking for every street in terms of crashes per foot
  mutate(index = 1:nrow(crashesperstreet)) %>%
  # We filter out Kelly Ave because it is an outlier - Experience says that there is no reason to think that Kelly Ave has that high of a ratio of crashes per foot of street, so I assume there must be an error in data entry or something like that
  filter(ST_FULL_NM != "SW KELLY AVE")
```
## Visualizations Relating to the Number of Crashes per Foot of Street
```{r}
# Graph of how the number of crashes per foot decreases in order from greatest to least
ggplot(crashesperft, aes(x = index, y = crashes_per_foot)) +
  geom_point(size = 0.75) +
  theme_bw() +
  ggtitle("Crashes per Foot, Ordered Highest to Lowest") +
  xlab("Street (Indexed Highest Crashes per Foot to Lowest)") +
  ylab("Crashes per Foot")
```

### Basic Modeling
```{r}
# Attempting to create a transformation (1 / X) in order to create a linear regression model
inversecrashesmodel <- lm(crashes_per_foot ~ I(1/(index)), data = crashesperft)
summary(inversecrashesmodel)

# Visualizing the transformation
ggplot(crashesperft, aes(x = (1/index), y = crashes_per_foot)) +
  geom_point(size = 0.75) +
  theme_bw() +
  ggtitle("Crashes per Foot with an x = (1 / Index) Transformation") +
  xlab("Street (1 / Index)") +
  ylab("Crashes per Foot")
```

### Playing with R-squared
I noticed that the R-squared value of 0.84 was pretty good, but if I played around with the transformation a little bit, I was able to increase the R-squared value.
```{r}
# This will make a vector of R-squared values of 
rsqvector <- c()
for(i in 0:30) {
  rsq <- summary(lm(crashes_per_foot ~ I(1/(index + i)), data = crashesperft))$r.squared
  rsqvector <- append(rsqvector, rsq)
}

ggplot(data.frame(r2 = rsqvector, ind = 0:30), aes(x = ind, y = r2)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = 0:30) +
  theme_classic() +
  ggtitle("R-squared Values of Each Model as a Function of the Shift in the Index") +
  xlab("Shift of Index in Model (1 / (Index + x))") +
  ylab("R-squared Value of Model")

max(rsqvector)
which.max(rsqvector) - 1

ggplot(crashesperft, aes(x = (1/(index + 5)), y = crashes_per_foot)) +
  geom_point(size = 0.75) +
  theme_bw() +
  ggtitle("Crashes per Foot (Highest R-squared Index Transformation)") +
  xlab("Street (1 / (Index + 5))") +
  ylab("Crashes per Foot")
```

Noticing the interesting geometry of the shift of index in the model and the R-squared value of the model.  Honestly, I have no idea what sort of mathematics causes these shapes to be so geometric, or why it is 5 that maximizes the R-squared value at 0.950065.

```{r}
maxrsq <- lm(crashes_per_foot ~ I(1/(index + 5)), data = crashesperft)
summary(maxrsq)
```
I kind of ran out of time for the project to play with the data a little bit more, which is a bit disappointing, because there is so much cool stuff in this dataset, but I think I will keep playing with it to discover anything else.  I think the index transformation stuff is quite interesting.  According to this model which may or may not be super accurate, if you gave me a ranking of a particular street compared to the others, I could predict how many crashes occur over the length of the street.
